---
title: "Cyclistic - Cleaning & EDA"
output: html_notebook
---

This notebook document the data cleaning process for the Cyclistic Project.   

Analysed Period: 2022/07/01 - 2023/06/30  
Data Source: https://divvy-tripdata.s3.amazonaws.com/index.html  

If you are following along with the process in R notebook, visit the link above and download trip data of the Analysed Period and place them in the cyclystic/raw_data folder, and run the cyclystic/scripts/01_load_data.R. Then we are ready to beging.  

### 1. Data Preview

First, import the combined raw data "trips_combined.csv"
```{r}
trips <- read_csv("../data/interim/trips_combined.csv")
```
<br>

Let's see what our data looks like
```{r}
head(trips)
```
<br>

Now let's get NA count for each variable
```{r}
print(summarise_all(trips, ~sum(is.na(.))))
```
<br>

Check for duplicates in `ride_id`. See result below:
```{r}
print(any(duplicated(trips$ride_id)))
```
<br>


### 2. Initial Cleaning from Data Preview

Extract "hour", "month", "weekday", and "trip duration" from from `started_at` and `ended_at`
```{r}
trips <- trips %>% 
  mutate(
    hour = hour(started_at),
    weekday = wday(started_at, label = TRUE),
    month = month(started_at, label = TRUE),
    trip_duration = difftime(ended_at, started_at, units = "mins")
  )
```
<br>

station name and `member_causal` column have different capitalization, let's make relevant character columns lower case. 
```{r}
trips <- trips %>% 
  mutate(across(c(rideable_type, start_station_name, end_station_name), tolower))
```
<br>

Trim white spaces in character columns just in case
```{r}
trips <- trips %>% mutate(across(where(is.character), trimws))
```
<br>

Make `rideable_type` and `member_casual` factors
```{r}
trips$member_casual <- as.factor(trips$member_casual)
trips$rideable_type <- as.factor(trips$rideable_type)
```
<br> 

We will investigate these items in section 3. Exploratory Data Analysis.  
- Station related columns all have around 15% NAs, suggesting a pattern rather than random.    
- The consistency in the level of precision for coordinates related columns needs to be verified.
- Handling NAs in `end_lat` and `end_lng`

### 3. Variable Exploration
<br>

##### Rideable Types & Trip Duration

First, look at distribution
```{r, fig.width=8, fig.height=4}
trip_duration_distribution_plot <- ggplot(trips, aes(x = trip_duration)) +
  geom_histogram(aes(fill = rideable_type), bins = 50, alpha = 0.7) +
  facet_wrap(~rideable_type) +
  labs(title = "Distribution of Trip Duration by Rideable Type", x = "Trip Duration (mins)", y = "Count") +
  scale_y_continuous(labels = comma) +
  theme_minimal() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 7), 
    legend.key.size = unit(7, "pt"),
    legend.position = "bottom", 
    )

trip_distribution_by_usertype_plot <- ggplot(trips, aes(x = rideable_type, fill = member_casual)) +
  geom_bar(position = "dodge") +
  labs(title = "Count of Rideable Types by User Type - `member_causal`",
       x = "Rideable Type",
       y = "Count") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 0.5),
    text = element_text(size = 7), 
    legend.key.size = unit(7, "pt"),
    legend.position = "bottom", 
    )

trip_duration_distribution_plot + trip_distribution_by_usertype_plot
```

Basic statistically analysis:
```{r}
trip_summary <- trips %>%
  group_by(rideable_type) %>%
  summarise(
    min_duration = min(trip_duration, na.rm = TRUE),
    max_duration = max(trip_duration, na.rm = TRUE),
    median_duration = median(trip_duration, na.rm = TRUE),
    average_duration = mean(trip_duration, na.rm = TRUE),
  )

# Printing results
print(trip_summary)
```

NOTES:
Negative trip duration is not possible but they are detected in both `classic_bike` and `eletric_bike`. Extreme long duration occurs in docked_bike which significantly skews the data. Both discoveries need further investigation. 

First, let's find out the percentage of negative duration trips exist in `classic_bike` and `eletric_bike`:
```{r}
neg_duration_rates <- trips %>%
  group_by(rideable_type) %>%
  summarise(rate = mean(trip_duration <= 0) * 100) %>%
  pivot_wider(names_from = rideable_type, values_from = rate)

print(glue("Percentage of negative duration in classic bike trips: {neg_duration_rates$classic_bike}%"))
print(glue("Percentage of negative duration in electric bike trips: {neg_duration_rates$electric_bike}%"))
```

NOTES:
The percentage of negative duration `classic_bike` and `eletric_bike` is considered insignificant. We'll treat them as noise and remove them from our dataset.

Remove `trips` with trips shorter than 2 min, negative duration trips included
```{r}
trips$trip_duration <- as.numeric(trips$trip_duration)

trips <- trips %>% 
  filter(trip_duration > 2)
```

<br> 

NOTE:
`docked_type` count is significantly lower than other 2 `rideable_types` and 100% of docked bikes are ridden by "casual" user type.
We will perform a bivariant analysis with just the "docked bike" `rideable_types` and `trip_duration` to gain a deeper understanding.

Log scale histogram
```{r, fig.width=8, fig.height=4}
ggplot(trips, aes(x = trip_duration)) +
  geom_histogram(aes(fill = rideable_type), bins = 50, alpha = 0.7) +
  facet_wrap(~rideable_type) +
  labs(title = "Distribution of Trip Duration by Rideable Type", x = "Trip Duration (mins)", y = "Count (Log Scale)") +
  scale_y_continuous(trans = "log10", labels = comma) +
  theme_minimal() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 7), 
    legend.key.size = unit(7, "pt"),
    legend.position = "bottom", 
  ) 
```
NOTE: We use log10 scale here to gain a deeper understanding of those abnormal long duration trips, ensuring they are not just rare random occurrences. The amount of abnormal long duration true plus its significantly lower ride count discovered earlier indicating a pattern rather than noise. Confirmed with further research, we've concluded that docked_bikes are bikes taken out of circulation, and should be removed for the data set for the purpose of this analysis.   

Remove `docked_bikes` trips from our analysis
```{r}
trips <- trips %>% 
  filter(rideable_type != "docked_bike")
```
<br>

NOTE: Note that both X and Y axis are on free scale, and still show a significant difference in the ride duration range between the two. Also, due to our skewed the data is, we are not going to use the outlier formula here. We'll use information provided by divvy trips website, and common sense to make our best judgement and focus on the core patterns. Based on the cost for each ride type, it's unlikely that a rider keeping a trip for long duration, ex: 5 hrs to 24 hrs on purpose since its get expensive. let's again use the log10 y scale to look at the 2 bike type's of trip duration distribution and spot patterns.  

```{r, fig.width=8, fig.height=4}
ggplot(trips, aes(x = trip_duration)) +
  geom_histogram(bins = 50, alpha = 0.7, color = "black") +
  facet_wrap(~rideable_type, scales = "free") +
  labs(title = "Distribution of Classic and Electric Bike Trip Duration", 
       x = "Trip Duration (mins)", 
       y = "Count") +
  scale_y_continuous(trans = "log10", labels = comma) +
  theme_minimal() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 7), 
    legend.key.size = unit(7, "pt"),
  )
```

NOTE: It's a log10 scale, so don't the graph miss lead you and pay attention to he 10 label on the y axis(count), We can see the first time a directional change is at around 531.25 min mark in the classic bike graph and 350 min mark which majority of the count are below a 100 at this point, and the trend start to flatten. We think this is attribute to bikes that were not docked properly earlier, either setting there waiting to be docked properly used again by other users, both resulted in the continuation of the duration, We also see a bike at the very end, this is likely due to the staff docking the the preciously not docking bike following schedule. We will remove trips longer than these two markers from our dataset. Though, they only represent a very small percentage of the dataset, our data should better refect the normal user behavior as a result. 

```{r}
trips <- trips %>%
  filter_and_report(
    (rideable_type == "classic_bike" & trip_duration < 531.25) |
    (rideable_type == "electric_bike" & trip_duration < 350)
  )
```

Finally, let's look at our distribution once again after it's been cleaned up 
```{r, fig.width=8, fig.height=4}
ggplot(trips, aes(x = trip_duration)) +
  geom_histogram(aes(fill = rideable_type), bins = 50, alpha = 0.7) +
  facet_wrap(~rideable_type) +
  labs(title = "Distribution of Trip Duration by Rideable Type", x = "Trip Duration (mins)", y = "Count (Log Scale)") +
  scale_y_continuous(labels = comma) +
  theme_minimal() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 7), 
    legend.key.size = unit(7, "pt"),
    legend.position = "bottom", 
  )
```
Trip Duration Summary
```{r}
trip_summary <- trips %>%
  group_by(rideable_type) %>%
  summarise(
    min_duration = min(trip_duration, na.rm = TRUE),
    max_duration = max(trip_duration, na.rm = TRUE),
    median_duration = median(trip_duration, na.rm = TRUE),
    average_duration = mean(trip_duration, na.rm = TRUE),
  )

# Printing results
print(trip_summary)
```


##### Explore NAs in Station Related Columns

Let's get an update on NAs in our dataset
```{r}
print(summarise_all(trips, ~sum(is.na(.))))
```

After further research, we learned that classic bikes dock exclusively at Divvy stations whereas electric bikes can dock at non-Divvy stations or public fixtures. This difference might explain the missing station ids and names in our data.Below, we will explore NAs in station columns of the different ride types.  

The Bar Char Represention
```{r, fig.width=8, fig.height=4}
missing_station <- trips %>% 
  filter(
    is.na(start_station_id) | 
    is.na(start_station_name) | 
    is.na(end_station_id) |
    is.na(end_station_name)
  )

ggplot(missing_station, aes(x = rideable_type)) +
  geom_bar(position = "dodge") +
  geom_text(
    aes(label = ..count..), 
    stat = 'count', 
    position = position_dodge(width = 0.9), 
    vjust = -0.25
  ) +
  labs(
    title = "Count of Missing Station Info",
    x = "Rideable Type",
    y = "NA Count"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 0.5),
    text = element_text(size = 7),  
    legend.key.size = unit(7, "pt"),  
    legend.position = "bottom"
  )
```
Our hypothesis is confirmed: the flexible return options for electric bikes are the primary cause of missing station IDs and names. These NA values are expected, therefore we will keep the rows with these NA values.  


##### NAs in End Coordinates

Now we can remove rows with NAs in end coordinates. See result below:
```{r}
trips <- trips %>% 
  filter_and_report(!is.na(end_lat) & !is.na(end_lng))
```

##### Imputation - Station Ids and Station Names
For both start and end, station ids come in pairs with station name, thus their count should be the same. There is a very small difference between them in our data set. We will use our imputation function to fix this. and print the missing value summary to see the difference
```{r}
trips <- impute_from_pair(trips, "start_station_name", "start_station_id")
trips <- impute_from_pair(trips, "end_station_name", "end_station_id")
print(summarise_all(trips, ~sum(is.na(.))))
```


##### Start & End Coordinates Precision Analysis
We can't improve the coordinate precision for those electric bike trips that either didn't start or end at a divvy station since there is no reference to a higher precision coordinates within our data. However, for trips that did start or end at a divvy bike station, through internal imputation, the accuracy can potentially be improved at scale. 

first, let's see how many divvy station actually has precise coordinates within our data. We will begin by adding  columns to label the precision in the form of decimal places, from 1 to 5. and get a count of unique stations with a precision of 5 decimal point or more vs total unique station exist in our data set
```{r}
trips <- trips %>% 
  mutate(
    start_coord_precision = get_max_precision_decimal(start_lat, start_lng),
    end_coord_precision = get_max_precision_decimal(end_lat, end_lng)
  )

low_precision_count <- (
  nrow(trips %>% filter(start_coord_precision < 5)) + 
    nrow(trips %>% filter(end_coord_precision < 5))
  )

# Create a reference of unique stations with precision coordinates
# ==============================================================================
# Below code needs update and refactoring achieve its purpose: 

# Create a subset for start stations and remove "start" prefix from selected columns
start_stations <- trips %>% 
  select(start_station_name, start_lat, start_lng, start_coord_precision) %>% 
  filter(!is.na(start_station_name)) %>% 
  rename(station_name = start_station_name,
         lat = start_lat,
         lng = start_lng,
         coord_precision = start_coord_precision)

# Create a subset for end stations and remove "end" prefix from all selected columns
end_stations <- trips %>% 
  select(end_station_name, end_lat, end_lng, end_coord_precision) %>% 
  filter(!is.na(end_station_name)) %>% 
  rename(station_name = end_station_name,
         lat = end_lat,
         lng = end_lng,
         coord_precision = end_coord_precision)

# Combine the two
stations_combined <- bind_rows(start_stations, end_stations)

# Create a subset of all unique stations regardless of coordinate precision 
stations_unique <- distinct(stations_combined, station_name)

# Keep only high precision coordinates (5 decimal places or higher)
# Keep only median coordinates value while removing duplicated stations
stations_precise <- stations_combined %>% 
  filter(coord_precision >= 5) %>% 
  group_by(station_name) %>% 
  summarise(med_lat = median(lat), med_lng = median(lng))

glue("Total unique stations: {nrow(stations_unique)}")
glue("Stations with precise coordinates: {nrow(stations_precise)}")
glue("Stations without precise coordinates: {nrow(stations_unique) - nrow(stations_precise)}")
glue("Total low precision coordinates count, both start and end: {low_precision_count}")

```

Before we move ahead with imputation, we must understand the potential impact and side effect of such operation. We will do so by investigating stations without precise coordinates.  

Let's first see how many trips are associated with these stations
```{r}
# Identify station without imprecise coordinates
stations_imprecise <- setdiff(
  stations_unique$station_name,
  stations_precise$station_name
)

# Subset the trips associated with these stations
trips_stations_imprecise <- trips %>%
  filter(
    start_station_name %in% stations_imprecise | 
     end_station_name %in% stations_imprecise
  )

glue("Trips associated with stations without precise coordinates: {nrow(trips_stations_imprecise)}")
```

The count is minimal, but we'll dive in a little deeper to spot any patterns.  
```{r}
ggplot(trips_stations_imprecise, aes(x = rideable_type, fill = member_casual)) +
  geom_bar(position = "dodge") +
  geom_text(
    aes(label = ..count..), 
    stat = 'count', 
    position = position_dodge(width = 0.9), 
    vjust = -0.25
  ) +
  labs(title = "Count of Rideable Types by User Type",
       x = "Rideable Type",
       y = "Count") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 0.5),
    text = element_text(size = 7), 
    legend.key.size = unit(7, "pt"),
    legend.position = "bottom", 
    )
```
All but one are electric bike rides. After view directly view these records, the only classic bike ride appears to have been returned to an incorrect station, given its end station name is "oh charging stx - test". This observation strengthens our hypothesis, especially since many station names have the prefix "public rack". The higher count of casual users here might be due to their unfamiliarity with the locations of divvy stations. 

Considered % of these records represent as of the total number of records, we feel it's sound to perform imputations. 

#### High Precision Coodindates Imputation 

```{r}
# Step 1: Join the main set with the reference set
trips <- trips %>%
  left_join(stations_precise, by = c("start_station_name" = "station_name")) %>%
  left_join(stations_precise, by = c("end_station_name" = "station_name"), suffix = c("_start", "_end"))

# Step 2: Impute coordinates with less than 5 decimal precision
trips <- trips %>%
  mutate(
    start_lat = ifelse(
      start_coord_precision < 5 & !is.na(med_lat_start), 
      add_randomness(med_lat_start),
      start_lat
    ),
    start_lng = ifelse(
      start_coord_precision < 5 & !is.na(med_lng_start),
      add_randomness(med_lng_start),
      start_lng
    ),
    end_lat = ifelse(
      end_coord_precision < 5 & !is.na(med_lat_end),
      add_randomness(med_lat_end),
      end_lat
    ),
    end_lng = ifelse(
      end_coord_precision < 5 & !is.na(med_lng_end),
      add_randomness(med_lng_end),
      end_lng
    )
  )

# Remove the extra columns that were created during the join if they are not needed
trips <- trips %>%
  select(-matches("_start$|_end$"))
```


Now let's re-impute the coordinate precision and compare before and after
```{r}
trips <- trips %>% 
  mutate(
    start_coord_precision = get_max_precision_decimal(start_lat, start_lng),
    end_coord_precision = get_max_precision_decimal(end_lat, end_lng)
  )

low_precision_count <- (
  nrow(trips %>% filter(start_coord_precision < 5)) + 
    nrow(trips %>% filter(end_coord_precision < 5))
  )

glue("Total low precision coordinates count, both start and end: {low_precision_count}")
```









